### 2021-12-16 documents/sections/extensions.tex
a thousand small fixes

\section{Inflections and verbs: innovation and resistance}
\label{sec:data}
As shown in \cref{sec:extensions_intro}, irregularly inflected first person forms are leftovers from incomplete person marker extensions.
\cref{sec:extensions} presents the six identified incomplete extensions, the prefixes they introduced and the verbs they spared.
Since the latter show considerable etymological overlap across languages, these resistant verbs are compared and reconstructed in \cref{sec:verbs}.
Where existent, their reflexes which did get affected by one of the extensions are identified.
%The main goal of this section is to identify the distributional patterns of innovative and conservative prefixes in the \gl{s_a_} lexicon, in order to search for potential explanations \pcref{sec:motivations}.

\subsection{Incomplete extensions: the innovative \gl{1}\texorpdfstring{\gl{s_a_}}{Sa} markers}
\label{sec:extensions}
As stated in \cref{sec:extensions_intro}, the six person marker extensions which did not affect all potential targets have in common that they introduced innovative first person markers on \gl{s_a_} verbs.
Of these extensions, half can be reconstructed to intermediate proto-languages, and half happened in pre-modern stages of single languages.
The sources of innovative markers vary, but not much: the innovative \gl{1}\gl{s_a_} prefix is formally identical to the \gl{1+2}\gl{p}/\gl{s_p_} marker (\PC \rc{k-}) in three cases, to the \gl{1}\gl{p}/\gl{s_p_} marker (\PC \rc{u(j)-}) in two cases, and to the \gl{1}>\gl{3} marker (\PC \rc{t-}) in one case.
For each extension, regular (innovative) verbs are contrasted with irregular (conservative) ones, and verb forms are reconstructed where necessary.
\cref{sec:pekodian} details the extension of \rc{k-} in \PPek, reflected in the three daughter languages \arara, \ikpeng, and \bakairi.
\cref{sec:waiwaian} treats the extension of \rc{k-} in \PWai, which was briefly shown in \cref{sec:extensions_intro}.
\cref{sec:taranoan} focuses on innovative \rc{t-} in \PTir, reflected in modern \trio and \akuriyo.
The topic of \crefrange{sec:akuriyo}{sec:yukpa} are innovative \gl{1}\gl{s_a_} markers only found in single languages:
\obj{k-} in \akuriyo, and \obj{j-} in \carijo and \yukpa.

@@

The most detailed description of a Pekodian language \parencite{alves2017arara} names six%
\footnote{Seven under her analysis, which sees the two meanings of \obj{it͡ʃi} \qu{to be, to lie down} as different verbs.}
\arara \gl{s_a_} verbs forming a subclass defined by a first person marker \obj{w(ɨ)-} rather than \obj{k-}, shown in \exref{arairr}.
There is also a reflex of the copula \rc{a[p]}, serving syntactically as a postposition introducing adverbial clauses meaning \qu{if} or \qu{when} \parencite[199--201]{alves2017arara}.
However, its inflectional morphology features verbal \setone prefixes, including first person \obj{w-} \exref{araap}.
%We take these \arara verbs as a starting point for the other Pekodian languages, as \textcite{alves2017arara} gives the most detailed description of person markers for any language of the branch.

@@

In his brief but precise discussion of \bakairi verbal person marking, \textcite{meira2003bakairi} reports the existence of two subclasses of \gl{s_a_} verbs, one taking first person \obj{w-}, and one \obj{k-}.\footnote{\textcite{meira2003bakairi} indicates that the same verbs which take first person \obj{w-} in \bakairi also take a \gl{1+2} marker \obj{k-}.
	However, this marker is only illustrated for \qu{to bathe}, both by \textcite{meira2003bakairi} and \textcite{von1892bakairi}.
	Given the lack of data for other verbs, this potential additional pattern will not be further discussed.
	If the characterization by \citeauthor{meira2003bakairi} is accurate, then verbs with innovative first person prefixes have conservative \gl{1+2} prefixes, and vice versa.
}
The first group is illustrated with \obj{i} \qu{to bathe} \exref{bak-33}.

\ex<bak-33>\bakairi \parencite[][4]{meira2003bakairi}\\
\begingl
\gla w-i-də//
\glb \gl{1}\gl{s_a_}-bathe-\gl{imm}//
\glft \qu{I bathed}//
\endgl
\xe
%
%Since \qu{to bathe} is also found in the \obj{w-}list for \arara, other \bakairi cognates of these verbs are of interest.
While \textcite[4]{meira2003bakairi} lists some \bakairi cognates of the \arara verbs in \exref{arairr} as \gl{s_a_} verbs, he does not indicate whether they belong to the \gl{s_a_}-1 class with \obj{k-}, or the \gl{s_a_}-2 class with \obj{w-}.
However, inflected forms can be found in \textcite{von1892bakairi}, presented in \exref{bakverbs} according to the analyses of \bakairi phonology and verbal morphology by \textcites{wheatley1969bakairi}{meira2003bakairi}{meira2005bakairi}{franchetto2016classes}.

@@

All descriptions of \ikpeng list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the verbs in question do not take \obj{k-}, but rather \obj{ɨ-} or Ø \exref{ikpw}, with the exception of \obj{k-}prefixed \qu{to go} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
While reflexes of \rc{a[p]} \qu{to be} do exist in \ikpeng, apparently only reflexes of \rc{eti} \qu{to be} occur with first person inflectional prefixes \parencite[401]{gildea2018reconstructing}.

@@

Reconstructions of verb stems are deferred to \cref{sec:verbs}, but a brief comment on \qu{to come} is in order:
The stems are not fully cognate, as \ikpeng and \bakairi both show a reflex of the \PPek detransitivizer \rc{əd-} in combination with a root reconstructible as \rc{epɨ}.
In contrast, the \arara first person form is directly based on this root \rc{epɨ}.
However, reflexes of \rc{əd-epɨ} can be found elsewhere in the \arara paradigm \exref{ara-123}.

\ex<ara-123>\arara \parencite[][150]{alves2017arara}\\
\begingl
\gla m-odebɨ-nɨ//
\glb \gl{2}\gl{s_a_}-come-\gl{rec}//
\glft \qu{You came.}//
\endgl
\xe
%
In contrast, \ikpeng and \bakairi show reflexes of \rc{əd-ebɨ} throughout the whole paradigm.
Following \posscite[114]{meira1998proto} line of reasoning for a similar pattern in Taranoan (see also \cref{sec:taranoan}), the idiosyncratic pattern in \arara can be reconstructed to \PPek, meaning that \bakairi and \ikpeng independently levelled the paradigm in favor of \rc{əd-epɨ}.

\subsubsection{\PWai \rc{k-}}
\label{sec:waiwaian}
This extension, one of the Parukotoan innovations shown in \cref{sec:extensions_intro}, resulted in the \hixka patterns from \cref{sec:intro}.
\PWai further extended the \gl{1}\gl{s_p_} prefix \rc{k-} innovated in \PPar to \gl{1}\gl{s_a_}.
For regularly inflected verbs, this created a unified \gl{1}\gl{s} category \pcref{tab:pwaireg}.
\input{floats/pwaireg}

@@

There are three more verbs which did not take innovative \rc{k-} in \PWai \pcref{tab:pwaiverbs}.
The two forms for \qu{to be} are unproblematic, whereas \qu{to go} is a special case.
While \hixka has the expected \obj{ɨ-}, \waiwai seems to have combined innovative \obj{k-} with the old \rc{w-}, an etymological analysis also considered by \textcite[90]{gildea1998}.
Alternatively, this form may have been influenced by deverbalized forms of \qu{to go}, where a reflex of the \gl{s_a_} class marker \rc{w-} has become fossilized \parentext{e.g., \obj{o-\emp{w}to-topo-nho} \qu{my trip} \parencite[92]{waiwaihawkins1998}}
%
%\pex<waiwto> \waiwai reflexes of the \gl{s_a_} class marker \rc{w-}
%\a \obj{o-\emp{w}to-topo-nho} \qu{my trip} \parencite[92]{waiwaihawkins1998}
%\a \obj{o-\emp{w}to-t͡ʃhe} \qu{after I went} \parencite[165]{waiwaihawkins1998}
%\a \obj{kɨ-\emp{w}to-me} \qu{for us to go} \parencite[204]{waiwaihawkins1998}
%\xe
%
In any case, the first person form \hixka \qu{to go} clearly points to \PWai \rc{wɨ-tom-}.

@@

\subsubsection{\PTir \rc{t-}}
\label{sec:taranoan}
The moniker Tiriyoan \parencite{glottolog} subsumes \trio and \akuriyo, the more closely related of the three Taranoan languages identified by \textcite{girard1971proto}, with \carijo as a more distant member.
\textcite{meira1998proto} contributes an extensive phonological, morphological, and lexical reconstruction of \PTar, facing an interesting puzzle in the \setone paradigms of \trio and \akuriyo: \PC \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} \rc{w-} seem to have switched places.
This resulted in a regular \gl{1}\gl{s_a_} marker of the form \envr{\rc{t͡ʃ-}}{\obj{e}}, \envr{\rc{t-}}{\obj{ə}} \pcref{tab:ptirreg}.%
\footnote{The latter allomorph was subsequently replaced by \obj{k-} in \akuriyo \pcref{sec:akuriyo}.}
\input{floats/ptirreg}
The question of how this switch happened in detail \parencite[107--112]{meira1998proto} still has no answer, although it seems necessary to assume a scenario in which both \rc{t-} and \rc{w-} for a time occurred on both transitive and intransitive verbs \parencite[112]{meira1998proto}.\footnote{
In fact, even the issue of \emph{when} this happened is open.
It could have happened at the \PTar stage, but the subsequent introduction of \obj{j-} in \carijo \pcref{sec:carijo} would have erased any traces of such an innovation.
Accordingly, \textcite{meira1998proto} hesitates to assign this extension to a specific proto-language.
Here, a conservative stance is taken and the innovation is arbitrarily assumed to be \PTir.
This decision does not affect the results of this study.}

@@

Turning to verbs not affected by the spread of \rc{t-}, \textcite{meira1998proto} reconstructs four of the items in \cref{tab:ptirverbs} as taking \rc{w-} in \PTar, for which reconstructed \PTir forms are substituted here.\footnote{The present reconstruction of \qu{to come} diverges from \posscite[114--115]{meira1998proto}, who reconstructs \PTar \rc{əepɨ} for first, but \rc{eepɨ} for other persons, based on the paradigmatic pattern in \trio and the vowel length in \akuriyo.
\akuriyo and \carijo would then have levelled that pattern, similar to what was suggested for the Pekodian languages \pcref{sec:pekodian}.
Here, the length in \akuriyo \obj{eepɨ} is identified as resulting from coalescence of \rc{əe}, and \trio \obj{əepɨ} as reflecting \rc{ətepɨ} \pcref{sec:come}.
This yields \PTir \rc{(əʔ)epɨ} and \PTar \rc{(ət͡ʃ-)epɨ}.}
The other copular root \rc{eʔi} can be added as a fifth verb, whose \trio reflex retains first person \obj{w-}.
The idiosyncratic \akuriyo first person prefix \obj{ə-} on \qu{to go} is identified as a reflex of \rc{wɨ-} by 
\textcite[113]{meira1998proto}, which is supported the fact that both components of the irregular change \rc{wɨ-} > \obj{ə-} (loss of \rc{w} and lowering of \rc{ɨ} to \obj{ə}) are found in other person prefixes \exref[exp.aku-164]{exp.aku-159}.

@@

\subsubsection{\akuriyo \obj{k-}}
\label{sec:akuriyo}
After the split of \PTir, when \rc{t-} and \rc{t͡ʃ-} had largely replaced \rc{w-}, \akuriyo innovated a third \gl{1}\gl{s_a_} marker \obj{k-}.
It seems to have replaced \rc{t-} only in specific environments, with the two markers showing a clear phonologically conditioned distribution in \posscite{gildea1994akuriyo} \akuriyo data \pcref{tab:aku1sa}.
\textcite[107]{meira1998proto} largely confirms that distribution, but mentions \dbqu{several cases of first person \obj{t-} in \akuriyo{}} (on \obj{ə}-initial verbs), albeit without any examples.
He also suggests that \obj{k-} might be more recent, which is plausible: since the distribution \envr{\rc{t-}}{\obj{ə}} / \envr{\rc{t͡ʃ-}}{\obj{e}} is reconstructible to \PTir, the most straightforward scenario is \obj{k-} replacing \rc{t-} but not \rc{t͡ʃ-} in \akuriyo.
The few \obj{t-} mentioned by \textcite{meira1998proto} were then perhaps reintroduced under \trio influence.
However, since there are no examples of, or more information about, \obj{ə}-initial verbs with \obj{t-}, these cases cannot be discussed further.

@@

\subsubsection{\carijo \obj{j-}}
\label{sec:carijo}
\carijo, the cousin of the Tiriyoan languages, has extended the \gl{1}\gl{s_p_} marker \obj{j(i)-}\footnote{Since all affected \gl{s_a_} verbs are V-initial, only the \envr{}{V} allomorph \obj{j-} occurs in that context.} to \gl{s_a_} verbs \parencite[105--107]{meira1998proto}.
Combined with the extension of \gl{2}\gl{s_a_} \obj{m-} and \gl{1+2}\gl{s_a_} \obj{kɨt-}/\obj{kɨs-} to \gl{s_p_} verbs, this created a single unified \gl{s} category for regular verbs \pcref{tab:carreg}.
%
\input{floats/carreg}%
%
Although the split-\gl{s} system has been lost entirely, former \gl{s_a_} verbs can be identified by the presence of a detransitivizer, like \obj{ehɨnəhɨ} \qu{to fight} \exref{car-30}, derived from \obj{hɨnəhɨ} \qu{to kill} \parencite[179]{robayo2000avance}.
%The same marker is also found in \gl{3}>\gl{1} scenarios \exref{car1p.car-1} and possessed nouns \exref{car1p.car-31}.

@@

As noted in \cref{sec:taranoan}, this extension also erased any traces of a putative \PTar \gl{1}\gl{s_a_} marker \rc{t-}.
However, it did not fully eclipse the old \gl{1}\gl{s_a_} marker \rc{w-}, which is attested as being preserved in the verbs \obj{tə} \qu{to go} \exref{car1w.car-24} and \obj{a} \qu{to be} \exref{car1w.car-25}.
In addition, the verb \obj{ka} \qu{to say} has a zero-marked first-person form \exref{car1w.car-33}.

\pex<car1w>\carijo \parencite[][5, 42, personal communication]{guerrero2016karihona}
\a<car-24>
\begingl
\gla wɨ-tə-e=rehe//
\glb \gl{1}-go-\gl{npst}=\gl{frust}//
\glft \qu{I almost go (but I am not going to go).}//
\endgl
\a<car-25>
\begingl
\gla əji-marə-ne w-a-e//
\glb \gl{2}-with-\gl{pl} \gl{1}-be-\gl{npst}//
\glft \qu{I am with you all.}//
\endgl
\a<car-33>
\begingl
\glpreamble dëmëmara kae ëwï iya//
\gla n-tə-mə=mara ∅-ka-e əwɨ i-ja//
\glb \gl{3}-go-\gl{pst}=\gl{dub} \gl{1}-say-\gl{npst}.\gl{cert} \gl{1}\gl{pro} \gl{3}-\gl{obl}//
\glft \qu{“Did s/he leave?”, I say to him.}//
\endgl
\xe
%We interpret this ∅ as a reflex of \rc{wɨ-}, a result of irregular phonological erosion.
Based on other C-initial verbs like \obj{tə} \qu{to go} or \obj{tuda} \qu{to arrive}, one would expect \obj{ka} \qu{to say} to either have conservative \obj{wɨ-} or innovative \obj{ji-}, so the zero needs explanation.
It is analysed here as a reflex of \rc{wɨ-}, primarily due to the loss of \rc{w} in \ikpeng and \hixka.
While those developments were more regular, an already irregular marker undergoing idiosyncratic phonological erosion is not that surprising, see \akuriyo \rc{wɨ-} > \obj{ə-} in the preceding section.
Alternatively, the divergent development of \rc{w-} on \obj{ka} \qu{to say} and \obj{tə} \qu{to go} may be due to the latter's originally V-initial nature \pcref{sec:go}.

\subsubsection{\yukpa \obj{j-}}
\label{sec:yukpa}
The divergent nature of the family-internal isolate \yukpa is \textit{inter alia} visible in the loss of many \setone forms and the formation of non-cognate innovative constructions \parencite{meira2006syntactic}.
However, it does preserve the \setone prefixes in the immediate past, which shows a unified intransitive paradigm \pcref{tab:yukreg}.
The wholesale loss of \gl{1+2} as an inflectional value was combined with the extension of \gl{2}\gl{s_a_} \obj{m(ɨ)-} to (now former) \gl{s_p_} verbs like \obj{nɨ} \qu{to sleep}.
%
\input{floats/yukreg}%
%
These verbs share their first person marker \obj{j(ɨ)-} with former \gl{s_a_} verbs like \obj{otum} \qu{to wash self}, identifiable by their semantics and the reflex of \detrz.
Since \obj{j(ɨ)-} is the reflex of the \PC \gl{1}S\textsubscript{(P)} marker \rc{u(j)-} \parencite[92]{gildea1998}, it also occurs on transitive verbs in \gl{3}>\gl{1} scenarios \exref{yukpatr.yuk-5}.
In contrast, \gl{1}>\gl{3} scenarios are zero-marked \exref{yukpatr.yuk-1}.

@@

In intransitive verbs, this first-person zero marking is attested in a single verb, \obj{to} \qu{to go} \exref{yuk-7}.
It diverges from regular C-initial verbs with \obj{jɨ-}, like \qu{to sleep} \pcref{tab:yukreg}.
It can thus be identified as having resisted the extension of \obj{j-} in \yukpa.

>>>>

\section{Prefixes and verbs: innovation and resistance}
\label{sec:data}
As shown in \cref{sec:extensions_intro}, irregularly inflected first person forms are leftovers of incomplete person marker extensions.
\cref{sec:extensions} presents the six identified incomplete extensions, the prefixes they introduced and the verbs they spared.
Since languages show considerable etymological overlap in their conservative verbs, they are compared and reconstructed in \cref{sec:verbs}.
Further, their reflexes which did get affected by an incomplete extension are identified.
%The main goal of this section is to identify the distributional patterns of innovative and conservative prefixes in the \gl{s_a_} lexicon, in order to search for potential explanations \pcref{sec:motivations}.

\subsection{Incomplete extensions: the innovative \gl{1}\texorpdfstring{\gl{s_a_}}{Sa} markers}
\label{sec:extensions}
As stated in \cref{sec:extensions_intro}, the six person marker extensions which left a group of verbs untouched all introduced innovative first person markers on \gl{s_a_} verbs.
Of these extensions, half can be reconstructed to intermediate proto-languages, and half happened in pre-modern stages of single languages.
The sources of innovative markers vary, but not much: the innovative \gl{1}\gl{s_a_} prefix is formally identical to the \gl{1+2}\gl{p}/\gl{s_p_} marker (\PC \rc{k-}) in three cases, to the \gl{1}\gl{p}/\gl{s_p_} marker (\PC \rc{u(j)-}) in two cases, and to the \gl{1}>\gl{3} marker (\PC \rc{t-}) in one case.
For each extension, regular (innovative) verbs are contrasted with irregular (conservative) ones, and verb forms are reconstructed where necessary.
\cref{sec:pekodian} demonstrates the extension of \rc{k-} in \PPek, reflected in the three daughter languages \arara, \ikpeng, and \bakairi.
\cref{sec:waiwaian} details the extension of \rc{k-} in \PWai, briefly shown in \cref{sec:extensions_intro}.
\cref{sec:taranoan} focuses on innovative \rc{t-} in \PTir, reflected in modern \trio and \akuriyo.
The topic of \crefrange{sec:akuriyo}{sec:yukpa} are innovative \gl{1}\gl{s_a_} markers found in single languages:
\obj{k-} in \akuriyo, and \obj{j-} in \carijo and \yukpa.

@@

The most detailed description of a Pekodian language \parencite{alves2017arara} names six%
\footnote{Seven in her analysis, treating the two meanings of \obj{it͡ʃi} \qu{to be, to lie down} as different verbs.}
\arara \gl{s_a_} verbs forming a subclass defined by a first person marker \obj{w(ɨ)-} rather than \obj{k-}, shown in \exref{arairr}.
There is also a reflex of the copula \rc{a[p]} \cref{sec:be}, serving syntactically as a postposition introducing adverbial clauses meaning \qu{if} or \qu{when} \parencite[199--201]{alves2017arara}.
However, it inflects with verbal \setone prefixes, including first person \obj{w-} \exref{araap}.
%We take these \arara verbs as a starting point for the other Pekodian languages, as \textcite{alves2017arara} gives the most detailed description of person markers for any language of the branch.

@@

In his brief but precise discussion of \bakairi verbal person marking, \textcite{meira2003bakairi} reports the existence of two \gl{s_a_} subclasses, defined by \gl{1}\gl{s_a_}\obj{w-} and \obj{k-}, respectively.\footnote{\textcite{meira2003bakairi} indicates that the same verbs which take first person \obj{w-} in \bakairi also take a \gl{1+2} marker \obj{k-}.
	However, this marker is only illustrated for \qu{to bathe}, both by \textcite{meira2003bakairi} and \textcite{von1892bakairi}.
	Given the lack of data for other verbs, this potential additional pattern will not be further discussed.
	If the characterization by \citeauthor{meira2003bakairi} is accurate, then verbs with innovative first person prefixes have conservative \gl{1+2} prefixes, and vice versa.
}
The first group is illustrated with \obj{i} \qu{to bathe} \exref{bak-33}.

\ex<bak-33>\bakairi \parencite[][4]{meira2003bakairi}\\
\begingl
\gla w-i-də//
\glb \gl{1}\gl{s_a_}-bathe-\gl{imm}//
\glft \qu{I bathed}//
\endgl
\xe
%
%Since \qu{to bathe} is also found in the \obj{w-}list for \arara, other \bakairi cognates of these verbs are of interest.
While \textcite[4]{meira2003bakairi} lists some \bakairi cognates of the \arara verbs in \exref{arairr} as \gl{s_a_} verbs, he does not indicate whether they belong to the \gl{s_a_}-1 class with \obj{k-}, or the \gl{s_a_}-2 class with \obj{w-}.
However, inflected forms can be found in \textcite{von1892bakairi}, represented in \exref{bakverbs} based on the analyses of \bakairi phonology and verbal morphology by \textcites{wheatley1969bakairi}{meira2003bakairi}{meira2005bakairi}{franchetto2016classes}.

@@

All descriptions of \ikpeng list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the verbs in question do not take \obj{k-}, but rather \obj{ɨ-} or Ø \exref{ikpw}, with the exception of \obj{k-}prefixed \qu{to go} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
While reflexes of \rc{a[p]} \qu{be-1} do exist in \ikpeng, apparently only reflexes of \rc{eti} \qu{be-2} occur with first person prefixes \parencite[401]{gildea2018reconstructing}.

@@

Reconstructions of verb stems are deferred to \cref{sec:verbs}, but a comment on \qu{to come} is in order:
The stems are not fully cognate, as \ikpeng and \bakairi both show a reflex of the \PPek detransitivizer \rc{əd-} in combination with a root reconstructible as \rc{epɨ} \pcref{sec:come}.
In contrast, the \arara first person form has a bare reflex of \rc{epɨ}.
While reflexes of \rc{əd-epɨ} can be found elsewhere in the \arara paradigm \exref{ara-123}, \ikpeng and \bakairi uniformly reflect \rc{əd-ebɨ}.

\ex<ara-123>\arara \parencite[][150]{alves2017arara}\\
\begingl
\gla m-odebɨ-nɨ//
\glb \gl{2}\gl{s_a_}-come-\gl{rec}//
\glft \qu{You came.}//
\endgl
\xe
%
Following \posscite[114]{meira1998proto} line of reasoning for a similar pattern in Taranoan (see also \cref{sec:taranoan}), the idiosyncratic pattern in \arara can be reconstructed to \PPek, with \bakairi and \ikpeng independently leveling the paradigm in favor of \rc{əd-epɨ}.

\subsubsection{\PWai \rc{k-}}
\label{sec:waiwaian}
This extension, one of the Parukotoan innovations shown in \cref{sec:extensions_intro}, resulted in the \hixka patterns from \cref{sec:intro}.
\PWai further extended the \gl{1}\gl{s_p_} prefix \rc{k-} (innovated in \PPar) to \gl{1}\gl{s_a_}.
For regularly inflected verbs, this created a unified \gl{1}\gl{s} category \pcref{tab:pwaireg}.
\input{floats/pwaireg}

@@

There are three more verbs which did not take innovative \rc{k-} in \PWai \pcref{tab:pwaiverbs}.
The two forms for \qu{to be} are unproblematic, whereas \qu{to go} is a special case.
While \hixka has the expected \obj{ɨ-}, \waiwai seems to have combined innovative \obj{k-} with  old \rc{w-}, an analysis also considered by \textcite[90]{gildea1998}.
Alternatively, this form may have been influenced by deverbalized forms of \qu{to go}, which show a fossilized of the \gl{s_a_} class marker \rc{w-}  \parentext{e.g., \obj{o-\emp{w}to-topo-nho} \qu{my trip} \parencite[92]{waiwaihawkins1998}}.
%
%\pex<waiwto> \waiwai reflexes of the \gl{s_a_} class marker \rc{w-}
%\a \obj{o-\emp{w}to-topo-nho} \qu{my trip} \parencite[92]{waiwaihawkins1998}
%\a \obj{o-\emp{w}to-t͡ʃhe} \qu{after I went} \parencite[165]{waiwaihawkins1998}
%\a \obj{kɨ-\emp{w}to-me} \qu{for us to go} \parencite[204]{waiwaihawkins1998}
%\xe
%
Either way, the first person form \hixka \qu{to go} clearly points to \PWai \rc{wɨ-tom-}.

@@

\subsubsection{\PTir \rc{t-}}
\label{sec:taranoan}
The moniker Tiriyoan \parencite{glottolog} subsumes \trio and \akuriyo, the more closely related of the three Taranoan languages identified by \textcite{girard1971proto}, with \carijo as a sister.
\textcite{meira1998proto} provides an extensive phonological, morphological, and lexical reconstruction of \PTar, facing an interesting puzzle in the \setone paradigms of \trio and \akuriyo: \PC \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} \rc{w-} seem to have switched places, creating a regular \gl{1}\gl{s_a_} marker of the form \envr{\rc{t͡ʃ-}}{\obj{e}}, \envr{\rc{t-}}{\obj{ə}} \pcref{tab:ptirreg}.%
\footnote{The latter allomorph was subsequently replaced by \obj{k-} in \akuriyo \pcref{sec:akuriyo}.}
\input{floats/ptirreg}
The question of how this switch happened in detail \parencite[107--112]{meira1998proto} is still open, although a scenario seems necessary in which both \rc{t-} and \rc{w-} for a time occurred on both transitive and intransitive verbs \parencite[112]{meira1998proto}.\footnote{
In fact, even the issue of \emph{when} this happened is open.
It could have happened at the \PTar stage, but the subsequent introduction of \obj{j-} in \carijo \pcref{sec:carijo} would have erased any traces of such an innovation.
Accordingly, \textcite{meira1998proto} hesitates to assign this extension to a specific proto-language.
Here, a conservative stance is taken and the innovation is arbitrarily assumed to be \PTir.
This decision does not affect the results of this study.}

@@

As for verbs unaffected by the spread of \rc{t-}, \textcite{meira1998proto} reconstructs four of the items in \cref{tab:ptirverbs} as taking \rc{w-} in \PTar, for which reconstructed \PTir forms are substituted here.\footnote{The present reconstruction of \qu{to come} diverges from \posscite[114--115]{meira1998proto}, who reconstructs \PTar \rc{əepɨ} for first, but \rc{eepɨ} for other persons, based on the paradigmatic pattern in \trio and the vowel length in \akuriyo.
\akuriyo and \carijo would then have levelled that pattern, similar to what was suggested for the Pekodian languages \pcref{sec:pekodian}.
Here, the length in \akuriyo \obj{eepɨ} is identified as resulting from coalescence of \rc{əe}, and \trio \obj{əepɨ} as reflecting \rc{ətepɨ} \pcref{sec:come}.
This yields \PTir \rc{(əʔ)epɨ} and \PTar \rc{(ət͡ʃ-)epɨ}.}
As a fifth verb \rc{eʔi} \qu{be-1} can be added, whose \trio reflex retains \obj{w-}.
The idiosyncratic \akuriyo first person prefix \obj{ə-} on \qu{to go} is identified as a reflex of \rc{wɨ-} by 
\textcite[113]{meira1998proto}, which is supported the fact that both components of the idiosyncratic change \rc{wɨ-} > \obj{ə-} (\phon{\rc{w}}{∅} and \phon{\rc{ɨ}}{\obj{ə}}) are found in other person prefixes \exref[exp.aku-164]{exp.aku-159}.

@@

\subsubsection{\akuriyo \obj{k-}}
\label{sec:akuriyo}
After the split of \PTir, when \rc{t-} and \rc{t͡ʃ-} had largely replaced \rc{w-}, \akuriyo innovated a third \gl{1}\gl{s_a_} marker \obj{k-}.
It seems to have replaced \rc{t-} only in specific environments, with \obj{k-} and \obj{t͡ʃ-} showing a clear phonologically conditioned distribution in \posscite{gildea1994akuriyo} \akuriyo data \pcref{tab:aku1sa}.
\textcite[107]{meira1998proto} largely confirms that distribution, but mentions \dbqu{several cases of first person \obj{t-} in \akuriyo{}} (on \obj{ə}-initial verbs), albeit without any examples.
He also suggests that \obj{k-} may be more recent, which is plausible: since the distribution \envr{\rc{t-}}{\obj{ə}} / \envr{\rc{t͡ʃ-}}{\obj{e}} is reconstructible to \PTir, the most straightforward scenario is \obj{k-} replacing \rc{t-} but not \rc{t͡ʃ-} in \akuriyo.
The few \obj{t-} mentioned by \textcite{meira1998proto} were then perhaps reintroduced under \trio influence.
However, since there are no examples of -- or more information about -- \obj{ə}-initial verbs with \obj{t-}, these cases cannot be discussed further.

@@

\subsubsection{\carijo \obj{j-}}
\label{sec:carijo}
\carijo, the third Taranoan language, extended the \gl{1}\gl{s_p_} marker \obj{j(i)-}\footnote{Since all affected \gl{s_a_} verbs are V-initial, only the \envr{}{V} allomorph \obj{j-} occurs in that context.} to \gl{s_a_} verbs \parencite[105--107]{meira1998proto}.
Combined with the extension of \gl{2}\gl{s_a_} \obj{m-} and \gl{1+2}\gl{s_a_} \obj{kɨt-}/\obj{kɨs-} to \gl{s_p_} verbs, this created a single unified \gl{s} category for regular verbs \pcref{tab:carreg}.
%
\input{floats/carreg}%
%
Although the split-\gl{s} system was lost entirely, former \gl{s_a_} verbs can be identified by the presence of a detransitivizer, like \obj{ehɨnəhɨ} \qu{to fight} \exref{car-30}, derived from \obj{hɨnəhɨ} \qu{to kill} \parencite[179]{robayo2000avance}.
%The same marker is also found in \gl{3}>\gl{1} scenarios \exref{car1p.car-1} and possessed nouns \exref{car1p.car-31}.

@@

As noted in \cref{sec:taranoan}, this extension also erased any traces of putative \PTar \gl{1}\gl{s_a_} \rc{t-}.
However, it did not fully eclipse the old \gl{1}\gl{s_a_} marker \rc{w-}, which is attested as being preserved in the verbs \obj{tə} \qu{to go} \exref{car1w.car-24} and \obj{a} \qu{to be} \exref{car1w.car-25}.
In addition, the verb \obj{ka} \qu{to say} has a zero-marked first-person form \exref{car1w.car-33}.

\pex<car1w>\carijo \parencite[][5, 42, personal communication]{guerrero2016karihona}
\a<car-24>
\begingl
\gla wɨ-tə-e=rehe//
\glb \gl{1}-go-\gl{npst}=\gl{frust}//
\glft \qu{I almost go (but I am not going to go).}//
\endgl
\a<car-25>
\begingl
\gla əji-marə-ne w-a-e//
\glb \gl{2}-with-\gl{pl} \gl{1}-be-\gl{npst}//
\glft \qu{I am with you all.}//
\endgl
\a<car-33>
\begingl
\glpreamble dëmëmara kae ëwï iya//
\gla n-tə-mə=mara ∅-ka-e əwɨ i-ja//
\glb \gl{3}-go-\gl{pst}=\gl{dub} \gl{1}-say-\gl{npst}.\gl{cert} \gl{1}\gl{pro} \gl{3}-\gl{obl}//
\glft \qu{“Did s/he leave?”, I say to him.}//
\endgl
\xe
%We interpret this ∅ as a reflex of \rc{wɨ-}, a result of irregular phonological erosion.
Based on other C-initial verbs like \obj{tə} \qu{to go} or \obj{tuda} \qu{to arrive}, \obj{ka} \qu{to say} should either have conservative \obj{wɨ-} or innovative \obj{ji-}, so the zero is unexpected.
It is analysed here as a reflex of \rc{wɨ-}, primarily due to the loss of \rc{w} in \ikpeng and \hixka.
While those developments were more regular, an already irregular marker undergoing idiosyncratic phonological erosion is not that surprising, see \akuriyo \rc{wɨ-} > \obj{ə-} in the preceding section.
Alternatively, the divergent development of \rc{w-} on \obj{ka} \qu{to say} and \obj{tə} \qu{to go} may be due to the latter's originally V-initial nature \pcref{sec:go}.

\subsubsection{\yukpa \obj{j-}}
\label{sec:yukpa}
The divergent nature of \yukpa\footnote{Very little is known about its sister \japreria, whose status as a dialect of \yukpa is contested by \textcite{oquendo2004japreria}.} is \textit{inter alia}t visible in the loss of many \setone forms and the formation of non-cognate innovative constructions \parencite{meira2006syntactic}.
However, it does preserve the \setone prefixes in the immediate past, which shows a unified intransitive paradigm \pcref{tab:yukreg}.
The wholesale loss of \gl{1+2} as an inflectional value was combined with the extension of \gl{2}\gl{s_a_} \obj{m(ɨ)-} to (now former) \gl{s_p_} verbs like \obj{nɨ} \qu{to sleep}.
%
\input{floats/yukreg}%
%
These verbs share their first person marker \obj{j(ɨ)-} with former \gl{s_a_} verbs like \obj{otum} \qu{to wash self}, identifiable by their semantics and the reflex of \detrz.
Since \obj{j(ɨ)-} is the reflex of the \PC \gl{1}S\textsubscript{(P)} marker \rc{u(j)-} \parencite[92]{gildea1998}, it also occurs on transitive verbs in \gl{3}>\gl{1} scenarios \exref{yukpatr.yuk-5}.
In contrast, \gl{1}>\gl{3} scenarios are zero-marked \exref{yukpatr.yuk-1}.

@@

In intransitive verbs, this first-person zero marking is attested with a single verb, \obj{to} \qu{to go} \exref{yuk-7}.
It diverges from regular C-initial verbs with \obj{jɨ-}, like \qu{to sleep} \pcref{tab:yukreg}.
It can thus be identified as having resisted the extension of \obj{j-} in \yukpa.
